---
title: "Building historical maps with CShapes"
author: "Bernhard Bieri"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building historical maps with CShapes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
pkgs <- c("WDI", "ggplot2", "dplyr", "countrycode", "qData", "qStates")
lapply(pkgs, library, character.only = TRUE)
```

Giving your readers a bird's eye view of your analysis can be a neat way to
convey complex concepts and observations of the structure of the international
governance system in a straightforward way. While not aiming at being a
comprehensive guide to the
[different ways of creating maps in R](https://geocompr.robinlovelace.net/),
this articles aims at introducing the process of building maps in R with data
from the excellent [`{cshapes}` package](https://cran.r-project.org/web/packages/cshapes/index.html)
complemented by GDP data by the `{WDI}` package which retrieves data from the
World Bank.

In the following example, we'll see how to build a heatmap showing GDP per
capita in 1980.

## Step 1: Import the data from `{cshapes}` and the World Bank

Below, we extract data from the `{WDI}` package of the
[NY.GDP.PCAP.KD](https://data.worldbank.org/indicator/NY.GDP.PCAP.KD) indicator
of the World Bank which yields information about the GDP per capita of countries
in 2010 $. We then clean this data and then convert the ISO country code to
the COW number we have in the `{cshapes}` dataset with the `{countrycode}`
package.

Furthermore, we leverage the `{cshapes}` package by importing a world map
in a way that is consistent with the qVerse with our `import_cshapes()`
function.

Finally, we perform a `left_join()` between both dataframes to combine mapping
data (especially the `geometry` variable which contains the shapes of countries)
and data about GDP in 1980.

```{r, echo=TRUE}
# Extract GDP from World Bank for the period between 1980 and 1990
library(WDI)
library(countrycode)
library(dplyr)

GDP <- WDI::WDI(indicator='NY.GDP.PCAP.KD', start=1980, end=1980)
GDP <- dplyr::slice(GDP, 50:266)
GDP <- dplyr::as_tibble(GDP)%>%
  qData::transmutate(COW_Nr = as.character(countrycode::countrycode(GDP$iso2c,
                                                       origin = "iso2c",
                                                       destination = "cown",
                                                       warn = FALSE)))

# Import map data with import_cshapes()
mapdata <- qStates::import_cshapes(date = "1980-01-01")

# Left join on COW_Nr
data <- dplyr::left_join(mapdata, GDP, by = "COW_Nr")
```

## Step 2: Create a heatmap

Below, we leverage `{ggplot}` and its `geom_sf()` layer that takes a vector of
polygons as an input (in this case the `geometry` variable of the combined
dataframe) to create a map. We also use a neat theme from the `{hrbrthemes}`
package and tweak it a little for a prettier output.

```{r, echo = TRUE}
library(ggplot2)
library(hrbrthemes) # for a pretty map

heatmap <- ggplot2::ggplot(data = data$geometry) + # Main plot options
  geom_sf(mapping = aes(fill = data$NY.GDP.PCAP.KD), lwd = 0.1) + # Map options
  scale_fill_viridis_c(option = "plasma", na.value = "grey80") + # Color options
  labs(title="GDP per capita in 1980",
       subtitle="Some fact about the GDP in 1980.",
       caption= "Source: World Bank (retrieved 15/07/2021)",
       fill = "GDP per capita") + # Define labels
  hrbrthemes::theme_ipsum() + # Set a default theme
  theme(panel.background = element_rect(fill = "light blue"), # Make oceans blue
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank()) # Remove grids
# Display the map
heatmap
```
