---
title: "Working with manystates data"
author: "Henrique Sposito and Jael Tan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with manystates data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      warning = FALSE, message = FALSE)
```

# Have democracies become more free and accountable since the 1990s?

`{manystates}` provides data on states in the international system across time,
assembling together datasets from reputable sources including
[The Correlates of War project](http://correlatesofwar.org),
[The Issue Correlates of War project](http://www.paulhensel.org/icow.html), and
[Gleditsch and Ward's dataset](https://doi.org/10.1080/03050629908434958) on interstate system membership.
The package can also be used to study related features such as
political regimes (`regimes` datacube), contiguity (`contiguity` datacube),
leadership (`leaders` datacube), and economic indicators (`economics` datacube).

To illustrate the uses of the data, we explore here the question of whether
states have indeed become more free and accountable.
Since 2017, more than half of the states in the world have become democracies.
However, at the same time,
influential works have theorized about the recent "democratic decay" and "stretch";
especially with the recent(ish) elections of leaders
with authoritarian tendencies in several democracies.

Below, we use `{manystates}` to study democratic accountability using data
from the `regimes` datacube as well as the `import_vdem()` function to extract 
[Variaties of Democracy Data (VDEM)](https://www.v-dem.net/project.html) data.
The `regimes` datacube contains data on political regimes in the Polity5 and Freedom House datasets.

## Data

For many years, Freedom House data has been widely relied upon,
including in the media, as an indicator of states that are democratic.
VDEM data has been developed more recently by scholars for a detailed and
comprehensive understanding of democracy.
In this vignette, we will focus on Freedom House and VDEM data
which include electoral, judicial, and civil dimensions.
Yet, Freedom House and VDEM projects also have slightly different approaches to the
measurement of democracy, which we will explore below.
Let's import the data first.
Due to the size of VDEM data,
we rely on the `import_vdem()` function to download the data,
contrary to the Freedom House dataset which is stored internally. 

```{r import}
library(manystates)
fh <- manystates::regimes$FreedomHouseFull
vdem <- import_vdem()
```

Let's examine first the data in VDEM by selecting a few variables
that are helpful to investigate freedom and accountability in democracies since the 1990s.

First, we need to know if a state is democratic and to what extent.
The "v2x_regime" variable in VDEM
differentiates between two types of democratic regimes (electoral and liberal) :

- Electoral democracy: free and fair multiparty elections...
- Liberal democracy: free and fair multiparty elections, access to justice, transparent law enforcement, respect for personal liberties, rule of law...

Second, we need indicators of democracy.
Let's focus on transparency and accountability ("v2x_accountability_osp"),
freedom of expression ("v2x_freexp_altinf"),
and civil society participation (v2x_cspart).

These variables were transformed into scales (e.g. 0-1) from
ordinal data from the surveys (ordered categorical) as follows:

- Accountability index: low to high (0-1)
- Freedom of expression: low to high (0-1)
- Civil society participation: low to high (0-1)

```{r subset vdem}
library(dplyr)
library(tidyr)
vdem <- vdem %>% dplyr::select(StateName, stateID, Year,
                              v2x_regime, v2x_accountability_osp,
                              v2x_freexp_altinf, v2x_cspart) %>%
  # select only the variables of interest
  dplyr::filter(Year > 1989, v2x_regime == 2 | v2x_regime == 3) %>%
  # filter for last 30 years (bigger than 1989) and for democracies only
  dplyr::rename(regime = v2x_regime,
                freedom_expression = v2x_freexp_altinf,
                accountability = v2x_accountability_osp,
                civil_society = v2x_cspart) %>% # rename some of these variables
  dplyr::mutate(regime_type = case_when(regime == 2 ~ "Electoral Democracy",
                                        regime == 3 ~ "Liberal Democracy")) %>% 
  # create a new regime type variable and rename the categories within
  arrange(desc(Year)) %>% # arrange by year
  drop_na() # drop rows where there is at least one NA obs...
```

The Freedom House data contains state year data starting from the year 2012.
Therefore, it is much smaller then VDEM.
Freedom House also organises its data somewhat differently,
disaggregating the concept of democracy into two components:
political rights ("PR rating" - "A" + "B" + "C") and
civil liberties ("CL rating" - "D" + "E" + "F" + "G").
The `FreedomHouseFull` dataset also contains the raw scores of each state for each question,
which are then grouped into categories labelled with alphabets from "A" to "G".
In terms of indicators of democracy, accountability is captured in category "C",
freedom of expression in category "D",
and civil society participation in category "B".
Let's subset the data accordingly.

```{r subset fh}
fh <- dplyr::select(fh, c(StateName, stateID, Year, PR, CL, B, C, D)) %>%
  # select only the variables of interest
  dplyr::filter(PR > 17) %>% # filter for democracies only
  manydata::transmutate(freedom_expression = D/16,
                        accountability = C/12,
                        civil_society = B/16) %>%
  # rename and standardise some of these variables
  dplyr::mutate(regime_type = ifelse(CL > 52, "Liberal Democracy",
                                     "Electoral Democracy")) %>%
  # create a new regime type variable and rename the categories within,
  # benchmark score derived from Scores table in Freedom House Codebook
  arrange(desc(Year)) %>% # arrange by year
  drop_na() # drop rows where there is at least one NA obs...
```

## Investigate the data

Let's first compare the 10 states with the highest accountability scores
(on average) in each dataset since 2012.

```{r investigate}
library(kableExtra)

vdem_account_10 <- vdem %>% # call data without assigning it
  dplyr::filter(Year > 2011) %>% # filter for years in both datasets
  group_by(StateName) %>% # group by state name
  summarise(Mean_VDEM = mean(accountability)) %>%
  arrange(desc(Mean_VDEM)) %>% # arrange the data by mean, in decreasing order
  slice(1:10)

fh_account_10 <- fh %>% # call data without assigning it
  group_by(StateName) %>% # group by state name
  summarise(Mean_FH = mean(accountability)) %>%
  # summarize grouped accountability variable by mean
  arrange(desc(Mean_FH)) %>%
  slice(1:10) # arrange the data by mean, in decreasing order

full_join(vdem_account_10, fh_account_10) %>% 
  kbl() %>% 
  kable_styling("striped")
```

It seems that the datasets differ somewhat on the states with highest accountability scores (on average). 

## Pathways: from electoral to liberal democracies (and vice-versa)

Let's compare which states appear as having multiple democratic regimes since 2012 in both datasets.

```{r transitions}
vdem_transition <- vdem %>% # saving this object to use later
  dplyr::filter(Year > 2011) %>% # filter for years in both datasets
  group_by(StateName, regime_type) %>% # group by state and regime
  count() %>% # count grouped observations
  group_by(StateName) %>% # re-group by state name only 
  mutate(Duplicated = n()) %>% # add duplicated columns for state names
  filter(Duplicated > 1) %>%  # filter for names that appear more than once
  select(-c(Duplicated)) %>% # remove duplicated variables
  rename('Years VDEM' = n, 'Regime Type' = regime_type) # rename for clarity

fh_transition <- fh %>% # saving this object to use later
  group_by(StateName, regime_type) %>% # group by state and regime
  count() %>% # count grouped observations
  group_by(StateName) %>% # re-group by state name only 
  mutate(Duplicated = n()) %>% # add duplicated columns for state names
  filter(Duplicated > 1) %>%  # filter for names that appear more than once
  select(-c(Duplicated)) %>% # remove duplicated variables
  rename('Years FH' = n, 'Regime Type' = regime_type) # rename for clarity

dplyr::full_join(vdem_transition, fh_transition) %>%
  arrange(StateName) %>%
  kbl() %>% 
  kable_styling("striped")
```

VDEM and freedom house appear to differ quite a bit in the cases that transitioned from liberal democracies to electoral democracies (and vice-versa) since 2012. Let's investigate further by taking a closer look at cases that appear in both datasets.

```{r}
transition_vdem <- vdem %>% # saving this object to use later
  dplyr::filter(Year > 2011) %>% # filter for years in both datasets
  group_by(StateName, regime_type) %>% # group by state and regime
  summarise(Years_VDEM = paste(Year, collapse = ", ")) %>%
  group_by(StateName) %>% # re-group by state name only 
  mutate(Duplicated = n()) %>% # add duplicated columns for state names
  filter(Duplicated > 1) %>%  # filter for names that appear more than once
  select(-c(Duplicated)) # remove duplicated variables

transition_fh <- fh %>% # saving this object to use later
  group_by(StateName, regime_type) %>% # group by state and regime
  summarise(Years_FH = paste(Year, collapse = ", ")) %>%
  group_by(StateName) %>% # re-group by state name only 
  mutate(Duplicated = n()) %>% # add duplicated columns for state names
  filter(Duplicated > 1) %>%  # filter for names that appear more than once
  select(-c(Duplicated)) # remove duplicated variables

inner_join(transition_vdem, transition_fh) %>% 
  kbl() %>%
  kable_styling("striped")
```

Even in the few cases of transtions from liberal democracies into electoral democracies (and vice-versa) since 2012 appear in both datasets, they seem to somewhat disagree in the specific years where these states are classified  as liberal and/or electoral democracies. 

Since VDEM has more state year observations and we are interested in democracy pathways since the 1990s, let's work with VDEM data only from now onwards.

Let's just look at the average differences for our democracy indicators for 
the states with multiple regimes.

```{r differences}
library(ggplot2)

vdem %>%
  group_by(StateName, regime_type) %>% # group by state and regime
  summarise(Accountability = mean(accountability),
            Freedom_of_Expression = mean(freedom_expression),
            Civil_Society = mean(civil_society)) %>%
  # summarize various grouped variable by mean
  group_by(StateName) %>% # re-group by state name only 
  mutate(Duplicated = n()) %>% # add duplicated columns for state names
  filter(Duplicated > 1) %>% # filter for names that appear more than once
  pivot_wider(names_from = regime_type,
              values_from = Accountability:Civil_Society,
              names_sep = "-") %>% # make data wide
  mutate(Accountability_dif = `Accountability-Liberal Democracy` -
           `Accountability-Electoral Democracy`,
         Freedom_of_Expression_dif = `Freedom_of_Expression-Liberal Democracy` -
           `Freedom_of_Expression-Electoral Democracy`,
         Civil_Society_dif = `Civil_Society-Liberal Democracy` -
           `Civil_Society-Electoral Democracy`) %>% 
  # create variables with mean difference
  select(StateName, Accountability_dif, Freedom_of_Expression_dif,
         Civil_Society_dif) %>% # select only pertinent observations
  pivot_longer(cols = Accountability_dif:Civil_Society_dif) %>% 
  # pivot table longer again
  rename(State = StateName,
         Indicator_Difference = name,
         Mean_Difference = value) %>% # rename for clarity
  ggplot(aes(x = Mean_Difference, y = State,
             fill = Indicator_Difference)) + # plot
  geom_point(aes(color = Indicator_Difference)) +
  facet_wrap("Indicator_Difference") +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "From Electoral to Liberal Democracies",
       y = "State",
       x = "Mean Difference")
```

Although the plot is not very informative, 
it appears that some states averages for these indicators were worse when
they were classified as liberal democracies in comparison to when they were
electoral democracies. This is puzzling.

Let's try a different approach, plotting averages for all indicators together
in time and by state.

```{r pathways}
vdem %>% # saving this object to use later
  group_by(StateName, regime_type) %>% # group by state and regime
  count() %>% # count grouped observations
  group_by(StateName) %>% # re-group by state name only 
  mutate(Duplicated = n()) %>% # add duplicated columns for state names
  filter(Duplicated > 1) %>%  # filter for names that appear more than once
  select(-c(Duplicated)) %>% # remove duplicated variables
  select(-n) %>% # remove years
  inner_join(vdem, by = c("StateName", "regime_type"), multiple = "all") %>%
  # keep only obs in present in both datasets
  mutate(Aggregated_Mean_Indicators = (accountability +
           freedom_expression + civil_society)/3) %>%
  # add rows to form the aggregated indicator
  select(StateName, regime_type, Year, Aggregated_Mean_Indicators) %>%
  # select only a few variables
  rename(State = StateName, Regime = regime_type) %>%
  ggplot(aes(x = as.numeric(Year), y = Aggregated_Mean_Indicators,
             colour = (Regime == "Liberal Democracy"))) + 
  geom_line(aes(group=1)) +
  facet_wrap("State") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(title = "From Electoral to Liberal Democracies (and vice-versa)",
       y = "Aggregated Mean Indicators") +
  scale_color_discrete(name="Regime",
                       labels=c("Electoral Democracy",
                                "Liberal Democracy"))
```

Once again, it appears that for some states, aggregated indicators for accountability,
freedom of expression, and civil society participation got better as they moved
from electoral democracy to liberal democracy (e.g. Taiwan),
or worse as they moved from liberal democracy to electoral democracy (e.g. Hungary). 
In other cases, it is not clear these changed at all (e.g. Benin) or,
even more puzzling, aggregated indicators got worse as they moved from electoral
democracy to liberal democracy (e.g. Greece),
or better as they moved from liberal democracy to electoral democracy (e.g. Slovakia).

# Indicators by region

Perhaps, however, we are missing important regional information that mediates the
relationship between these indicators and whether states are classified 
liberal or electoral democracies.
How about we check the evolution of these indicators by region?
Let's get and join region data from the `HUGGO_STATES` datacube to see the
evolution of the aggregated indicators per region since the 1990s.

```{r regions}
HUGGO_STATES <- states$HUGGO_STATES %>%
  select(StateName, Region) %>% # select only the variables we want
  distinct() # keeps only distinct rolls (no duplicates)

vdem %>%
  dplyr::left_join(HUGGO_STATES, by = "StateName") %>%
  group_by(Region) %>%
  mutate(max_region_year = max(as.numeric(Year)),
         min_region_year = min(as.numeric(Year))) %>% # get min and max values for each regions
  filter(Year == min_region_year | Year == max_region_year) %>%
  # filter for first and last year of data
  group_by(Region, Year) %>% # group region and year
  summarise(Aggregated_Mean_Indicators = (mean(accountability) +
                                            mean(freedom_expression) +
                                            mean(civil_society))/3) %>% 
  # add aggregated means by groups
  drop_na() %>%
  ggplot(aes(x = Aggregated_Mean_Indicators,
      y = reorder(Region, Aggregated_Mean_Indicators),
      Fill = Region)) +
  geom_point(aes(color = Region)) +
  geom_line(aes(color = Region)) +
  geom_text(aes(label = Year), nudge_y = 0.3) +
  theme_minimal() +
  theme(legend.position="none") +
  labs(title = "Are democracies better off in 2021 than in 1990?",
       subtitle = "Difference aggregated means for indicators of
       accountability, freedom of expression, and civil society indicators",
       y = "Region",
       x = "Mean Difference")
```

Perhaps a better question is **where** have democracies become more free and
accountable since the 1990s.
